---
title: "HW3"
author: "Zian ZHUANG"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(KMsurv)
library(survival)
library(tidyverse)
```

## 4.2
Using the data reported in section 1.3, find the quantities specified
below for the AML low risk and AML high risk groups. Note that most
of these quantities are worked out in detail in Example 4.2 and its
continuations for the ALL group.

### a
Estimate the survival functions and their standard errors for the AML
low risk and AML high risk groups.

```{r}
data(bmt)

#low risk group
fit_low <- survfit(Surv(t2, d3) ~ 1, data = bmt[which(bmt$group==2),])
summary(fit_low)

#high risk group
fit_high <- survfit(Surv(t2, d3) ~ 1, data = bmt[which(bmt$group==3),])
summary(fit_high)
```

### b
Estimate the cumulative hazard rates and their standard errors for the AML low risk and AML high risk groups.

```{r}
# low risk group
H.est <- cumsum(fit_low$n.event/fit_low$n.risk)
Var.est <- cumsum(fit_low$n.event/fit_low$n.risk^2)
data.frame(H.est=H.est, SE.est=sqrt(Var.est))

# high risk group
H.est <- cumsum(fit_high$n.event/fit_high$n.risk)
Var.est <- cumsum(fit_high$n.event/fit_high$n.risk^2)
data.frame(H.est=H.est, SE.est=sqrt(Var.est))
```

### d

Estimate the mean time to death and find 95% confidence intervals for the mean survival time for both the AML low risk and AML high risk groups. (Answers are given in section 4.5.)

```{r}
#try hand write equation to calculate for low risk group
ind <- which(bmt$group==2&bmt$d3==1&bmt$t2<2081)
low_day <- bmt$t2[ind] %>% sort()
dif_low_day <- c(low_day[1], diff(c(low_day,2081)))
sum(dif_low_day * c(1,summary(fit_low)$sur[-25]))

ind <- which(bmt$group==3&bmt$d3==1&bmt$t2<2081)
high_day <- bmt$t2[ind] %>% sort()
dif_high_day <- c(high_day[1], diff(c(high_day,2081)))
dif_high_day <- dif_high_day[which(dif_high_day!=0)]
sum(dif_high_day * c(1,summary(fit_high)$sur))

# low risk group
#mean se
print(fit_low, rmean=2081)
#CI
c(1315-1.96*119, 1315+1.96*119)

high_day <- fit_high$time
dif_high_day <- c(high_day[1], diff(high_day))
sum(dif_high_day * fit_high$surv)

# high risk group
#mean se
print(fit_high, rmean=2081)
#CI
c(656-1.96*123, 656+1.96*123)

```

### e

Work out estimates of the median time to death and find 95% confidence intervals for the median survival time for both the AML low risk and AML high risk groups using the linear, log-transformed, and arcsine formulas. (Answers are given in section 4.5.)

```{r}
source("functions//bc_median_ci.R")

# low risk group
bc.median.ci(fit_low, type = "linear", table = F)
bc.median.ci(fit_low, type = "log", table = F)
bc.median.ci(fit_low, type = "asin", table = F)

# high risk group
bc.median.ci(fit_high, type = "linear", table = F)
bc.median.ci(fit_high, type = "log", table = F)
bc.median.ci(fit_high, type = "asin", table = F)
```

### f

Find 95% confidence intervals for the survival functions at 300 days post-transplant for both the AML low risk and AML high risk groups using the log- and arcsine-transformed formulas.


```{r}
source("functions//arcsin_ci.R")
```

#### Using arcsine-transformed

AML low risk:
```{r}
res <- arcsin.ci(fit_low, alpha = 0.05)
res
```

S(300)=S(288), then we can found that 95% confidence intervals for the survival functions is 0.778(0.658, 0.878).

AML high risk:
```{r}
res <- arcsin.ci(fit_high, alpha = 0.05)
res
```

S(300)=S(273), then we can found that 95% confidence intervals for the survival functions is 0.422(0.283, 0.568).

#### Using log-transformed formulas.

Firstly, define a function to calculate

```{r}
log.ci <- function(surv.object, alpha = 0.05, events.only = FALSE) {
  
  #- Extract necessary values from survsurv.object object
  surv.est  <- surv.object$surv
  sigma     <- surv.object$std.err #This is sigma, not the SE of S(t)
  surv.time <- surv.object$time
  
  #- Get limits element wise
  lims <- data.frame(time = surv.object$time, surv = surv.est,  sigma = surv.object$std.err)
  
  theta <- exp(qnorm(1 - alpha / 2) * sigma / log(surv.est))
  lims$lower <- surv.est^(1/theta)
  lims$upper <- surv.est^(theta)
  
  if (events.only == TRUE) {
    return(round(lims[which(surv.object$n.censor == 0), ], 3)) 
  } else {
    return(round(lims, 3))
  }
}
```

For AML low risk:

```{r}
log.ci(fit_low, alpha = 0.05)
```

S(300)=S(288), then we can found that 95% confidence intervals for the survival functions is 0.778(0.642, 0.867).

For AML high risk:

```{r}
log.ci(fit_high, alpha = 0.05)
```

S(300)=S(273), then we can found that 95% confidence intervals for the survival functions is 0.422(0.278, 0.560).

#### likelihood ratio method

For AML low risk:
```{r}
source("functions//LRci.R")
LRci.surv(fit_low, t = 300)
```

For AML high risk:
```{r}
LRci.surv(fit_high, t = 300)
```

#### bootstrap method

```{r}
source("functions//SurvivalProb_boots.R")
Low_group <- bmt[which(bmt$group==2),]
High_group <- bmt[which(bmt$group==3),]
```

For AML low risk:

```{r}
SurvivalProb_boots(Low_group$t2, Low_group$d3, 300, plots = TRUE)
```

Then we can calculate the 95%CI

```{r}
c(0.7777778-1.96*0.05521416, 0.7777778+1.96*0.05521416)
```

For AML high risk:

```{r}
SurvivalProb_boots(High_group$t2, High_group$d3, 300, plots = TRUE)
```

Then we can calculate the 95%CI

```{r}
c(0.4222222-1.96*0.07486193, 0.4222222+1.96*0.07486193)
```


### g
Find 95% EP confidence bands for the survival functions over the range 100–400 days post-transplant for both the AML low risk and AML high risk groups using the linear, log-transformed, and arcsinetransformed formulas.

```{r}
source("functions\\conf_bands.R")
```


#### AML low risk


```{r}
#calculate critical regions first
cbands.region(fit_low, 100, 400)
```

linear
```{r}
cbands.interval(fit_low, 100, 400, crit.value=(2.6430+2.5949)/2, 
                alpha = 0.05, type = "linear", method = "ep")
```

arcsinetransformed
```{r}
cbands.interval(fit_low, 100, 400, crit.value=(2.6430+2.5949)/2, 
                alpha = 0.05, type = "asin", method = "ep")
```

log-transformed
```{r}
cbands.interval(fit_low, 100, 400, crit.value=(2.6430+2.5949)/2, 
                alpha = 0.05, type = "log", method = "ep")
```

#### AML high risk


```{r}
#calculate critical regions first
cbands.region(fit_high, 100, 400)
2.6998+2.6811
```

linear
```{r}
cbands.interval(fit_high, 100, 400, crit.value=(2.6998+2.6811)/2, 
                alpha = 0.05, type = "linear", method = "ep")
```

arcsinetransformed
```{r}
cbands.interval(fit_high, 100, 400, crit.value=(2.6998+2.6811)/2, 
                alpha = 0.05, type = "asin", method = "ep")
```

log-transformed
```{r}
cbands.interval(fit_high, 100, 400, crit.value=(2.6998+2.6811)/2, 
                alpha = 0.05, type = "log", method = "ep")
```

### h
Find 95% HW confidence bands for the survival functions over
the range 100–400 days post-transplant for both the AML low risk and
AML high risk groups using the linear, log-transformed, and arcsinetransformed formulas.


#### AML low risk


```{r}
#calculate critical regions first
cbands.region(fit_low, 100, 400)
```

linear
```{r}
cbands.interval(fit_low, 100, 400, crit.value=(1.0294+1.0281)/2, 
                alpha = 0.05, type = "linear", method = "hw")
```

arcsinetransformed
```{r}
cbands.interval(fit_low, 100, 400, crit.value=(1.0294+1.0281)/2, 
                alpha = 0.05, type = "asin", method = "hw")
```

log-transformed
```{r}
cbands.interval(fit_low, 100, 400, crit.value=(1.0294+1.0281)/2, 
                alpha = 0.05, type = "log", method = "hw")
```

#### AML high risk

```{r}
#calculate critical regions first
cbands.region(fit_high, 100, 400)
```

linear
```{r}
cbands.interval(fit_high, 100, 400, crit.value=(1.3201+1.3155)/2, 
                alpha = 0.05, type = "linear", method = "hw")
```

arcsinetransformed
```{r}
cbands.interval(fit_high, 100, 400, crit.value=(1.3201+1.3155)/2, 
                alpha = 0.05, type = "asin", method = "hw")
```

log-transformed
```{r}
cbands.interval(fit_high, 100, 400, crit.value=(1.3201+1.3155)/2, 
                alpha = 0.05, type = "log", method = "hw")
```


### i
Based on the results above and those discussed in Example 4.2 and its continuations, how do the survival experiences of the ALL, AML low risk, and AML high risk groups compare?

```{r}
survdiff(Surv(t2, d3) ~ group, data=bmt)
```


As we can tell from the results above and those discussed in Example 4.2 and its continuations, AML low risk has the highest estimated mean&median survival probability and AML high risk has the lowest estimated mean&median survival probability. All estimate methods provided consistent answers. In addition, the difference is significant according to the Log-Rank Test. 

## 4.7
Consider a hypothetical study of the mortality experience of diabetics. Thirty diabetic subjects are recruited at a clinic and followed until death or the end of the study. The subject’s age at entry into the study and their age at the end of study or death are given in the table below. Of interest is estimating the survival curve for a 60- or for a 70-year-old diabetic

### a

Since the diabetics needed to survive long enough from birth until the study began, the data is left truncated. Construct a table showing the number of subjects at risk, Y , as a function of age.

```{r}
(df<-data.frame(time=c(58:74,76:80),
               Y=c(2,3,5,6,9,10,10,10,10,12,13,14,13,12,12,11,9,7,5,4,3,1)))
```


### b

Estimate the conditional survival function for the age of death of a diabetic patient who has survived to age 60.

```{r}
df %>% 
  mutate(n = c(2,3,4,6,8,9,10,8,9,12,11,12,11,10,10,10,8,6,4,4,3,1)) %>% 
  mutate(S_60 = cumprod(n/Y))
```


### c

Estimate the conditional survival function for the age of death of a diabetic patient who has survived to age 70.

```{r}
df %>% 
  mutate(n = c(2,3,4,6,8,9,10,8,9,12,11,12,11,10,10,10,8,6,4,4,3,1)) %>% 
  mutate(S_60 = cumprod(c(rep(1,12),(n/Y)[-c(1:12)])))
```

### d

Suppose an investigator incorrectly ignored the left truncation and simply treated the data as right censored. Repeat parts a–c

```{r}
(df<-data.frame(time=c(58:74,76:80),
                Y=c(rep(30,3),29,29,28,27,27,25,24,24,22,18,16,14,11,9,7,5,4,3,1)))
```

```{r}
df %>% 
  mutate(n=Y-c(0,0,1,0,1,1,0,2,1,0,2,2,2,2,2,1,1,1,1,0,0,0)) %>% 
  mutate(S_60 = cumprod(n/Y))
```

```{r}
df %>% 
  mutate(n=Y-c(0,0,1,0,1,1,0,2,1,0,2,2,2,2,2,1,1,1,1,0,0,0)) %>% 
  mutate(S_60 = cumprod(c(rep(1,12),(n/Y)[-c(1:12)])))
```

