---
title: "COVID-19 project report 2"
author: Zian ZHUANG
output:
  html_document:
    highlight: pygments
    toc: yes
    toc_float: yes
---

<!-- Setup -->
<style type="text/css">
body{ /* Normal  */
      font-size: 17px;
      font-family: "Times New Roman";
  }
h1,h2,h3,h4,h5,h6{
  font-family: "Times New Roman";
}
</style>

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.align = 'center', message = FALSE)
library(knitr)
library(KMsurv)
library(ggplot2)
library(survival)
library(tidyverse)
library(kableExtra)
```
<!-- Begin writing -->

## Q4.6

In section 1.6 a study is described to evaluate a protocol change in disinfectant practice in a large midwestern university medical center. Control of infection is the primary concern for the 155 patients entered into the burn unit with varying degrees of burns. The outcome variable is the time until infection from admission to the unit. Censoring variables are discharge from the hospital without an infection or death without an infection. Eighty-four patients were in the group which had chlorhexidine as the disinfectant and 72 patients received the routine disinfectant povidone-iodine.

### (a)
Estimate the survival (infection-free) functions and their standard errors for the chlorhexidine and povidone-iodine groups.

```{r}
data(burn)
df_p <- burn[which(burn$Z1==0),]
df_c <- burn[which(burn$Z1==1),]

#povidone-iodine group
fit_p <- survfit(Surv(T3, D3) ~ 1, data = df_p)
summary(fit_c)

#chlorhexidine group
fit_c <- survfit(Surv(T3, D3) ~ 1, data = df_c)
summary(fit_c)
```

### (b)
Estimate the cumulative hazard rates and their standard errors for the chlorhexidine and povidone-iodine groups. Plot these estimates. Does it appear that the two cumulative hazard rates are proportional to each other?

```{r}
# povidone-iodine group
H.est.p <- cumsum(fit_p$n.event/fit_p$n.risk)
Var.est.p <- cumsum(fit_p$n.event/fit_p$n.risk^2)
data.frame(H.est=H.est.p, SE.est=sqrt(Var.est.p))

# chlorhexidine group
H.est.c <- cumsum(fit_c$n.event/fit_c$n.risk)
Var.est.c <- cumsum(fit_c$n.event/fit_c$n.risk^2)
data.frame(H.est=H.est.c, SE.est=sqrt(Var.est.c))

# proportion
H.est.p/H.est.c
```


### (c)
Provide estimates of the median time to infection and find 95% confidence intervals for the median time to infection for both the chlorhexidine and povidone-iodine groups using the linear, log-transformed, and arcsine formulas.

```{r}
source("functions//bc_median_ci.R")

# povidone-iodine group
bc.median.ci(fit_p, type = "linear", table = F)
bc.median.ci(fit_p, type = "log", table = F)
bc.median.ci(fit_p, type = "asin", table = F)

# chlorhexidine group
bc.median.ci(fit_c, type = "linear", table = F)
bc.median.ci(fit_c, type = "log", table = F)
bc.median.ci(fit_c, type = "asin", table = F)
```

### (d)
Find 95% confidence intervals for the survival (infection-free) functions at 10 days postadmission for both the chlorhexidine and povidone-iodine groups using the log transformed and arcsine transformed formulas.

#### arcsin transformed

For povidone-iodine group
```{r}
source("functions//arcsin_ci.R")
arcsin.ci(fit_p, alpha = 0.05)
```
we can found that 95% confidence intervals for the survival functions at 10 days is 0.856(0.774, 0.923).

For chlorhexidine group
```{r}
arcsin.ci(fit_c, alpha = 0.05)
```
we can found that 95% confidence intervals for the survival functions at 10 days is 0.736(0.626, 0.834).

#### log transformed

Firstly, define a function to calculate

```{r}
log.ci <- function(surv.object, alpha = 0.05, events.only = FALSE) {
  
  #- Extract necessary values from survsurv.object object
  surv.est  <- surv.object$surv
  sigma     <- surv.object$std.err #This is sigma, not the SE of S(t)
  surv.time <- surv.object$time
  
  #- Get limits element wise
  lims <- data.frame(time = surv.object$time, surv = surv.est,  sigma = surv.object$std.err)
  
  theta <- exp(qnorm(1 - alpha / 2) * sigma / log(surv.est))
  lims$lower <- surv.est^(1/theta)
  lims$upper <- surv.est^(theta)
  
  if (events.only == TRUE) {
    return(round(lims[which(surv.object$n.censor == 0), ], 3)) 
  } else {
    return(round(lims, 3))
  }
}
```

For povidone-iodine group
```{r}
log.ci(fit_p, alpha = 0.05)
```
we can found that 95% confidence intervals for the survival functions at 10 days is 0.856(0.761, 0.916).

For chlorhexidine group
```{r}
log.ci(fit_c, alpha = 0.05)
```
we can found that 95% confidence intervals for the survival functions at 10 days is 0.736(0.614, 0.825).

### (e)
Find 95% confidence bands for the infection-free functions over the range 8–20 days postinfection for both the chlorhexidine and povidone-iodine groups using the linear, log transformed, and arcsine transformed formulas.

```{r}
source("functions\\conf_bands.R")
```


#### povidone-iodine

```{r}
#calculate critical regions first
cbands.region(fit_p, 8, 20)
#2.6299
```

linear
```{r}
cbands.interval(fit_p, 8, 20, crit.value=2.6299, 
                alpha = 0.05, type = "linear", method = "ep")
```

arcsinetransformed
```{r}
cbands.interval(fit_p, 8, 20, crit.value=2.6299, 
                alpha = 0.05, type = "asin", method = "ep")
```

log-transformed
```{r}
cbands.interval(fit_p, 8, 20, crit.value=2.6299, 
                alpha = 0.05, type = "log", method = "ep")
```

#### chlorhexidine

```{r}
#calculate critical regions first
cbands.region(fit_c, 8, 20)
#(2.5374+2.4859+2.5678+2.5193)/4
```

linear
```{r}
cbands.interval(fit_c, 8, 20, crit.value=(2.5374+2.4859+2.5678+2.5193)/4, 
                alpha = 0.05, type = "linear", method = "ep")
```

arcsinetransformed
```{r}
cbands.interval(fit_c, 8, 20, crit.value=(2.5374+2.4859+2.5678+2.5193)/4, 
                alpha = 0.05, type = "asin", method = "ep")
```

log-transformed
```{r}
cbands.interval(fit_c, 8, 20, crit.value=(2.5374+2.4859+2.5678+2.5193)/4, 
                alpha = 0.05, type = "log", method = "ep")
```

### (f)
Find 95% HW confidence bands for the infection-free functions over the range 8–20 days postinfection for both the chlorhexidine and povidone-iodine.

### (g)
Based on the results above, how does the infection experience of the chlorhexidine and povidone-iodine groups compare?

## Q4.7

Consider a hypothetical study of the mortality experience of diabetics. Thirty diabetic subjects are recruited at a clinic and followed until death or the end of the study. The subject’s age at entry into the study and their age at the end of study or death are given in the table below. Of interest is estimating the survival curve for a 60- or for a 70-year-old diabetic

### (a)

Since the diabetics needed to survive long enough from birth until the study began, the data is left truncated. Construct a table showing the number of subjects at risk, Y , as a function of age.

```{r}
(df<-data.frame(time=c(58:74,76:80),
               Y=c(2,3,5,6,9,10,10,10,10,12,13,14,13,12,12,11,9,7,5,4,3,1)))
```


### (b)

Estimate the conditional survival function for the age of death of a diabetic patient who has survived to age 60.

```{r}
df %>% 
  mutate(n = c(2,3,4,6,8,9,10,8,9,12,11,12,11,10,10,10,8,6,4,4,3,1)) %>% 
  mutate(S_60 = cumprod(n/Y))
```


### (c)

Estimate the conditional survival function for the age of death of a diabetic patient who has survived to age 70.

```{r}
df %>% 
  mutate(n = c(2,3,4,6,8,9,10,8,9,12,11,12,11,10,10,10,8,6,4,4,3,1)) %>% 
  mutate(S_60 = cumprod(c(rep(1,12),(n/Y)[-c(1:12)])))
```

### (d)

Suppose an investigator incorrectly ignored the left truncation and simply treated the data as right censored. Repeat parts a–c

```{r}
(df<-data.frame(time=c(58:74,76:80),
                Y=c(rep(30,3),29,29,28,27,27,25,24,24,22,18,16,14,11,9,7,5,4,3,1)))
```

```{r}
df %>% 
  mutate(n=Y-c(0,0,1,0,1,1,0,2,1,0,2,2,2,2,2,1,1,1,1,0,0,0)) %>% 
  mutate(S_60 = cumprod(n/Y))
```

```{r}
df %>% 
  mutate(n=Y-c(0,0,1,0,1,1,0,2,1,0,2,2,2,2,2,1,1,1,1,0,0,0)) %>% 
  mutate(S_60 = cumprod(c(rep(1,12),(n/Y)[-c(1:12)])))
```


## Q4.8

Table 1.7 reports the results of a study on the survival times of patients admitted to a psychiatric hospital. In this data set patients were admitted to the hospital at a random age and followed until death or the end of the study. Let X be the patient’s age at death. Note that the data we have on X is left truncated by the patient’s age at entry into the hospital and right censored by the end of the study.

### (a)
Plot the number at risk, Yi , as a function of age.

### (b)
Estimate the conditional survival function for a psychiatric patient who has survived to age 30 without entering a psychiatric hospital.

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>