---
title: 'Survival Analysis Using R'
output: 
  html_document:
    toc: yes
---
## Introduction

This lab serves as an introduction to using R for survival analysis. 

First we will need to load the required packages. 

```{r}
#install.packages(survival)
#install.packages(KMsurv)
library(survival)  # loading survival functions into R
library(KMsurv)    # datasets from Klein and Moeschberger textbook
```

We will be using datasets from Klein and Moeschberger's *Survival Analysis: Techniques for Censored and Truncated Data 2ed*. Let's load the breast cancer trial dataset from section 1.5.

```{r}
#After you load KMsurv
data(btrial)
help(btrial) # Info about the dataset.
head(btrial) # First 6 rows of the dataset.
```

time: Time to death or on-study time, months
death: Death indicator (0=alive, 1=dead)
im: Immunohistochemical response (1=negative, 2=positive)

## The Surv Object

Before we start fitting a model to the data, we must first create a ```Surv``` object in ```R```. Two inputs are needed to create this object, a vector of observed survival times, and a corresponding vector of censoring indicators. It is standard that uncensored events are labeled as "1", and censored events are "0" for the censoring indicators. The Surv object identifies to ```R``` which event times should be treated as censored. For the ```btrial``` dataset, it will be the time, death columns. To extract these, we can use the ```$``` symbol.

```{r}
help(Surv)
Surv(btrial$time, btrial$death) # What does this tell you?
```

##Fitting a Parametric Survival Regression Model 
We will discuss how to fit what are more commonly known as accelerated failure time models later in the course. A very brief introduction will be given here as a toy example.

Much like in linear regression where we need to assume normality, for parametric survival models we need to assume that the survival time follows a certain distribution. The most common assumptions are the exponential, and Weibulll. Others include, but are not limited to, log-logistic, log-normal, and generalized gamma. The survival function for the exponential distribution is
$$S(t) = \exp(-\lambda t), \hspace{.2in} t > 0, \lambda > 0.$$
For the Weibull distribution we have
$$S(t) = \exp\{-(\lambda t)^\alpha\} \hspace{.2in} t > 0, \lambda > 0, \alpha > 0.$$

Note that the exponential distribution is a special case of the Weibull distribution where $\alpha = 1$. To fit a parametric survival model we will use the ```survreg( )``` command from the survival package. The functionality is very similar to that of ```lm( )``` for linear regression. 

```{r}
#Fitting an intercept-only model (ie. No covariates)
help(survreg)

fit.exp <- survreg(Surv(btrial$time, btrial$death) ~ 1, dist = "exponential")

fit.wei <- survreg(Surv(btrial$time, btrial$death) ~ 1, dist = "weibull")
```

You can use the ```summary( )``` command to extract p-values, estimates, etc. 

**Question:** How can one test if an exponential models fits better than a Weibull model?

```{r}
#Results:
summary(fit.exp); summary(fit.wei)

ls(fit.exp) #What we can extract from this object
```

Note that we get estimates for the intercept and scale term. However, these aren't estimates of $\lambda$ and $\alpha$, which are generally of interest. There is a relationship however.

$$\hat{\lambda} = \exp(-intercept)$$
$$\hat{\alpha} = \frac{1}{scale}$$
Other distributional assumptions of observed survival times:
```{r}
fit.gau <- survreg(Surv(btrial$time, btrial$death) ~ 1, dist = "gaussian")
summary(fit.gau)
```
```{r}
fit.logit <- survreg(Surv(btrial$time, btrial$death) ~ 1, dist = "logistic")
summary(fit.logit)
```
```{r}
fit.lognorm <- survreg(Surv(btrial$time, btrial$death) ~ 1, dist = "lognormal")
summary(fit.lognorm)
```

**Question:** How can we estimate standard errors for $\hat{\lambda}$ and $\hat{\alpha}$?

###Adding Covariates 

We previously fit an intercept-only model. What if we were interested in adding covariates? We can add covariates similarly to how we would when using ```lm```. 

```{r}
## An expoential survival model
#Let’s re-parametrize im.
btrial$newim <- ifelse(btrial$im == 1, 1, 0)
survreg(Surv(time,death)~newim, data = btrial, dist = "exponential")
```

```{r}
## A weibull survival model
survreg(Surv(time,death)~newim, data = btrial, dist = "weibull")
```